\bf Visualising How Non-linear Dimension Reduction Warps Your Data

### 2024-03-27 paper.tex
fixed on the _ fig labels - this breaks in current version of quarto

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{agsm}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Visualising How Non-linear Dimension Reduction Warps Your Data},
  pdfauthor={Jayani P.G. Lakshika; Dianne Cook; Paul Harrison; Michael Lydeamore; Thiyanga S. Talagala},
  pdfkeywords={high-dimensional data, dimension
reduction, triangulation, hexagonal binning, low-dimensional
manifold, manifold learning, tour, data vizualization},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, sharp corners, frame hidden, breakable, enhanced, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt]}{\end{tcolorbox}}\fi

\hypertarget{sec-intro}{%
\section{Introduction}\label{sec-intro}}

@@

{\centering \includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-nldervisUMAP-1.pdf}

@@

\end{figure}

\hypertarget{sec-background}{%
\section{Background}\label{sec-background}}

\hypertarget{dimension-reduction}{%
\subsection{Dimension Reduction}\label{dimension-reduction}}

@@

\hypertarget{non-linear-dimension-reduction-techniques}{%
\subsubsection{Non-linear dimension reduction
techniques}\label{non-linear-dimension-reduction-techniques}}

@@

{\centering \includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-nldervis-1.pdf}

@@

\end{figure}

\hypertarget{linear-overviews-using-tours}{%
\subsection{Linear overviews using
tours}\label{linear-overviews-using-tours}}

@@

\hypertarget{sec-methods}{%
\section{Methodology}\label{sec-methods}}

@@

{\centering \includegraphics[width=1\textwidth,height=1\textheight]{figures/workflow.png}

@@

\end{figure}

\hypertarget{preprocessing-steps}{%
\subsection{Preprocessing steps}\label{preprocessing-steps}}

@@

\hypertarget{sec-construct2d}{%
\subsection{Constructing the 2D model}\label{sec-construct2d}}

@@

\begin{equation}\protect\hypertarget{eq-equation1}{}{
 ar = \frac{r_2}{r_1}
}\label{eq-equation1}\end{equation}

\begin{equation}\protect\hypertarget{eq-equation2}{}{
 hr = \frac{hb}{wb}
}\label{eq-equation2}\end{equation}

\begin{equation}\protect\hypertarget{eq-equation3}{}{
 y_{max} = \left\lceil\frac{ar}{hr}\right\rceil * hr
}\label{eq-equation3}\end{equation}

@@

\begin{equation}\protect\hypertarget{eq-equation11}{}{
 h = \sqrt{3} * s
}\label{eq-equation11}\end{equation}

\begin{equation}\protect\hypertarget{eq-equation12}{}{
 b_1 = \frac{r_1 + buffer_{x}}{h}
}\label{eq-equation12}\end{equation}

\begin{equation}\protect\hypertarget{eq-equation13}{}{
  v = 1.5 * s
}\label{eq-equation13}\end{equation}

\begin{equation}\protect\hypertarget{eq-equation14}{}{
 b_2 = \frac{r_2 + buffer_{y}}{v}
}\label{eq-equation14}\end{equation}

@@

\begin{equation}\protect\hypertarget{eq-equation15}{}{
d_x = \frac{h}{2}
}\label{eq-equation15}\end{equation}

\begin{equation}\protect\hypertarget{eq-equation16}{}{
d_y = \frac{2* v * 1.15}{\sqrt{3}} 
}\label{eq-equation16}\end{equation}

@@

{\centering \includegraphics[width=1\textwidth,height=1\textheight]{figures/hex_coord.png}

@@

\end{figure}

@@

\begin{equation}\protect\hypertarget{eq-equation4}{}{
C_k^{(2)} = (C_{ky_1}, C_{ky_2}) = \left(\frac{\sum_{i=1}^{6} h^{k}x_{i}}{6}, \frac{\sum_{i=1}^{6} h^{k}y_{i}}{6}\right).
}\label{eq-equation4}\end{equation}

@@

\begin{equation}\protect\hypertarget{eq-equation5}{}{
C_k^{(2)} = (C_{ky_1}, C_{ky_2}) = \left(\frac{1}{n_k} \sum_{i=1}^{n_k} y_{1i}, \frac{1}{n_k} \sum_{i=1}^{n_k} y_{2i}\right),
}\label{eq-equation5}\end{equation}

@@

\hypertarget{lifting-the-model-into-high-dimensions}{%
\subsection{Lifting the model into high
dimensions}\label{lifting-the-model-into-high-dimensions}}

\hypertarget{lifting-the-triangular-mesh-points-into-high-dimensions}{%
\subsubsection{Lifting the triangular mesh points into high
dimensions}\label{lifting-the-triangular-mesh-points-into-high-dimensions}}

@@

\begin{equation}\protect\hypertarget{eq-equation6}{}{
C_k^{(p)} = (C_{kx_1}, ..., C_{kx_p}) = \left(\frac{1}{n_k} \sum_{i=1}^{n_k} x_{1i}, \frac{1}{n_k} \sum_{i=1}^{n_k} x_{2i}, \dots ,\frac{1}{n_k} \sum_{i=1}^{n_k} x_{pi}\right).
}\label{eq-equation6}\end{equation}

@@

\begin{equation}\protect\hypertarget{eq-equation7}{}{
v(C_k^{(2)}) = C_k^{(p)}.
}\label{eq-equation7}\end{equation}

@@

\hypertarget{lifting-the-2d-triangular-mesh-into-high-dimensions}{%
\subsubsection{Lifting the 2D triangular mesh into high
dimensions}\label{lifting-the-2d-triangular-mesh-into-high-dimensions}}

@@

{\centering \includegraphics{paper_files/figure-pdf/fig-wkhighD-1.pdf}

@@

\end{figure}

\hypertarget{tuning-the-model}{%
\subsection{Tuning the model}\label{tuning-the-model}}

@@

\hypertarget{total-number-of-bins}{%
\subsubsection{Total number of bins}\label{total-number-of-bins}}

@@

{\centering \includegraphics{paper_files/figure-pdf/fig-binsize-1.pdf}

@@

\end{figure}

\begin{equation}\protect\hypertarget{eq-equation8}{}{
b = b_1 \times b_2
}\label{eq-equation8}\end{equation}

@@

{\centering \includegraphics{paper_files/figure-pdf/fig-diagnosticpltScurve-1.pdf}

@@

\end{figure}

\hypertarget{benchmark-value-to-remove-low-density-hexagons}{%
\subsubsection{Benchmark value to remove low-density
hexagons}\label{benchmark-value-to-remove-low-density-hexagons}}

@@

\begin{equation}\protect\hypertarget{eq-equationp2}{}{
\text{standard count} = \frac{\text{count}}{\text{max count}} 
}\label{eq-equationp2}\end{equation}

@@

{\centering \includegraphics{paper_files/figure-pdf/fig-stdctsScurve-1.pdf}

@@

{\centering \includegraphics{paper_files/figure-pdf/fig-diagnosticpltScurvelwd-1.pdf}

@@

{\centering \includegraphics{paper_files/figure-pdf/fig-rmlowdenshex-1.pdf}

@@

\end{figure}

\hypertarget{benchmark-value-for-removing-long-edges}{%
\subsubsection{Benchmark value for removing long
edges}\label{benchmark-value-for-removing-long-edges}}

@@

{\centering \includegraphics{paper_files/figure-pdf/fig-distScurve-1.pdf}

@@

{\centering \includegraphics{paper_files/figure-pdf/fig-modelScurve-1.pdf}

@@

\end{figure}

\hypertarget{starting-point-of-the-hexagonal-grid}{%
\subsubsection{Starting point of the hexagonal
grid}\label{starting-point-of-the-hexagonal-grid}}

@@

{\centering \includegraphics{paper_files/figure-pdf/fig-mseScurvestartpoint-1.pdf}

@@

{\centering \includegraphics{paper_files/figure-pdf/fig-scurveshifthexgridsexp-1.pdf}

@@

{\centering \includegraphics{paper_files/figure-pdf/fig-scurveshifthexgrids-1.pdf}

@@

\end{figure}

\hypertarget{sec-summary}{%
\subsection{Model summaries}\label{sec-summary}}

\hypertarget{predicted-values-and-residuals}{%
\subsubsection{Predicted values and
residuals}\label{predicted-values-and-residuals}}

@@

\begin{equation}\protect\hypertarget{eq-equation10}{}{
\text{Error} = \sum_{j = 1}^{n}\sum_{i = 1}^{p} (x_{ij} - C_{x_ij})^2
}\label{eq-equation10}\end{equation}

@@

\hypertarget{sec-goodfit}{%
\subsubsection{Goodness of fit statistics}\label{sec-goodfit}}

@@

\begin{equation}\protect\hypertarget{eq-equation9}{}{
\text{MSE} = \sum_{j = 1}^{n} \frac{\sum_{i = 1}^{p} (x_{ij} - C_{x_ij})^2}{n}
}\label{eq-equation9}\end{equation}

\hypertarget{sec-simpleex}{%
\subsection{Simulated data example}\label{sec-simpleex}}

@@

{\centering \includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-nldervis5Gau-1.pdf}

@@

\end{figure}

To investigate which is the best representation to visualize the
original data or all NLDR methods provide equivalent information, we
visualize the model-in-data space. Models from all NLDR methods show
five well-separated clusters (see Figure~\ref{fig-gau1_sc},
Figure~\ref{fig-gau2_sc}, Figure~\ref{fig-gau3_sc},
Figure~\ref{fig-gau4_sc}, and Figure~\ref{fig-gau5_sc}). This suggests
that for the five Gaussian cluster dataset, all NLDR methods effectively
preserve the global structure. tSNE, UMAP, and PHATE display clusters
with varying densities, indicating their ability to capture
within-cluster variation (see Figure~\ref{fig-gau1_sc},
Figure~\ref{fig-gau2_sc}, and Figure~\ref{fig-gau3_sc}). On the other
hand, both TriMAP and PaCMAP show clusters with flat surfaces,
suggesting a failure to capture within-cluster variation (see
Figure~\ref{fig-gau4_sc} and Figure~\ref{fig-gau5_sc}). Therefore,
TriMAP and PaCMAP do not capture the local structure as effectively as
other methods.

@@

{\centering \includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-modelfiveGau-1.pdf}

@@

\begin{minipage}[t]{0.33\linewidth}

{\centering

\raisebox{-\height}{

\includegraphics{figures/five_gau_clusters/sc_tsne_1.png}

@@

@@

\subcaption{\label{fig-gau1_sc1}}
\end{minipage}%
%
\begin{minipage}[t]{0.33\linewidth}

{\centering

\raisebox{-\height}{

\includegraphics{figures/five_gau_clusters/sc_tsne_2.png}

@@

@@

\subcaption{\label{fig-gau1_sc2}}
\end{minipage}%
%
\begin{minipage}[t]{0.33\linewidth}

{\centering

\raisebox{-\height}{

\includegraphics{figures/five_gau_clusters/sc_tsne_3.png}

@@

@@

\subcaption{\label{fig-gau1_sc3}}
\end{minipage}%

\caption{\label{fig-gau1_sc}Screen shots of the \textbf{langevitour} of
the five Gaussian clusters dataset, shows the model with tSNE in high-D,
a video of the tour animation is available at
(\url{https://youtu.be/oQxEb4wRdHI}).}

\end{figure}

@@

\begin{minipage}[t]{0.33\linewidth}

{\centering

\raisebox{-\height}{

\includegraphics{figures/five_gau_clusters/sc_umap_1.png}

@@

@@

\subcaption{\label{fig-gau2_sc1}}
\end{minipage}%
%
\begin{minipage}[t]{0.33\linewidth}

{\centering

\raisebox{-\height}{

\includegraphics{figures/five_gau_clusters/sc_umap_2.png}

@@

@@

\subcaption{\label{fig-gau2_sc2}}
\end{minipage}%
%
\begin{minipage}[t]{0.33\linewidth}

{\centering

\raisebox{-\height}{

\includegraphics{figures/five_gau_clusters/sc_umap_3.png}

@@

@@

\subcaption{\label{fig-gau2_sc3}}
\end{minipage}%

\caption{\label{fig-gau2_sc}Screen shots of the \textbf{langevitour} of
the five Gaussian clusters dataset, shows the model with UMAP in high-D,
a video of the tour animation is available at
(\url{https://youtu.be/JW49csPpDx4}).}

\end{figure}

@@

\begin{minipage}[t]{0.33\linewidth}

{\centering

\raisebox{-\height}{

\includegraphics{figures/five_gau_clusters/sc_phate_1.png}

@@

@@

\subcaption{\label{fig-gau3_sc1}}
\end{minipage}%
%
\begin{minipage}[t]{0.33\linewidth}

{\centering

\raisebox{-\height}{

\includegraphics{figures/five_gau_clusters/sc_phate_2.png}

@@

\subcaption{\label{fig-gau3_sc2}}
\end{minipage}%
%
\begin{minipage}[t]{0.33\linewidth}

{\centering

\raisebox{-\height}{

\includegraphics{figures/five_gau_clusters/sc_phate_3.png}

\subcaption{\label{fig-gau3_sc3}}
\end{minipage}%

\caption{\label{fig-gau3_sc}Screen shots of the \textbf{langevitour} of
the five Gaussian clusters dataset, shows the model with PHATE in
high-D, a video of the tour animation is available at
(\url{https://youtu.be/xqhRODt8Z-o}).}

\end{figure}

\begin{minipage}[t]{0.33\linewidth}

{\centering

\raisebox{-\height}{

\includegraphics{figures/five_gau_clusters/sc_trimap_1.png}

\subcaption{\label{fig-gau4_sc1}}
\end{minipage}%
%
\begin{minipage}[t]{0.33\linewidth}

{\centering

\raisebox{-\height}{

\includegraphics{figures/five_gau_clusters/sc_trimap_2.png}

\subcaption{\label{fig-gau4_sc2}}
\end{minipage}%
%
\begin{minipage}[t]{0.33\linewidth}

{\centering

\raisebox{-\height}{

\includegraphics{figures/five_gau_clusters/sc_trimap_3.png}

\subcaption{\label{fig-gau4_sc3}}
\end{minipage}%

\caption{\label{fig-gau4_sc}Screen shots of the \textbf{langevitour} of
the five Gaussian clusters dataset, shows the model with TriMAP in
high-D, a video of the tour animation is available at
(\url{https://youtu.be/X-uHSB4VoZ0}).}

\end{figure}

\begin{minipage}[t]{0.33\linewidth}

{\centering

\raisebox{-\height}{

\includegraphics{figures/five_gau_clusters/sc_pacmap_1.png}

\subcaption{\label{fig-gau5_sc1}}
\end{minipage}%
%
\begin{minipage}[t]{0.33\linewidth}

{\centering

\raisebox{-\height}{

\includegraphics{figures/five_gau_clusters/sc_pacmap_2.png}

\subcaption{\label{fig-gau5_sc2}}
\end{minipage}%
%
\begin{minipage}[t]{0.33\linewidth}

{\centering

\raisebox{-\height}{

\includegraphics{figures/five_gau_clusters/sc_pacmap_3.png}

\subcaption{\label{fig-gau5_sc3}}
\end{minipage}%

\caption{\label{fig-gau5_sc}Screen shots of the \textbf{langevitour} of
the five Gaussian clusters dataset, shows the model with PaCMAP in
high-D, a video of the tour animation is available at
(\url{https://youtu.be/UxvU8HQd7JM}).}

\end{figure}

\hypertarget{sec-applications}{%
\section{Applications}\label{sec-applications}}

\hypertarget{single-cell-rna-seq-data-of-human}{%
\subsection{Single-cell RNA-seq data of
human}\label{single-cell-rna-seq-data-of-human}}

@@

{\centering \includegraphics{paper_files/figure-pdf/fig-umapauthor-1.pdf}

@@

\end{figure}

>>>>

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{agsm}
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Visualising How Non-linear Dimension Reduction Warps Your Data},
  pdfauthor={Jayani P.G. Lakshika; Dianne Cook; Paul Harrison; Michael Lydeamore; Thiyanga S. Talagala},
  pdfkeywords={high-dimensional data, dimension
reduction, triangulation, hexagonal binning, low-dimensional
manifold, manifold learning, tour, data vizualization},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\section{Introduction}\label{sec-intro}

@@

\centering{

\includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-nldervisUMAP-1.pdf}

@@

\end{figure}%

\section{Background}\label{sec-background}

\subsection{Dimension Reduction}\label{dimension-reduction}

@@

\subsubsection{Non-linear dimension reduction
techniques}\label{non-linear-dimension-reduction-techniques}

@@

\centering{

\includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-nldervis-1.pdf}

@@

\end{figure}%

\subsection{Linear overviews using
tours}\label{linear-overviews-using-tours}

@@

\section{Methodology}\label{sec-methods}

@@

\centering{

\includegraphics[width=1\textwidth,height=1\textheight]{figures/workflow.png}

@@

\end{figure}%

\subsection{Preprocessing steps}\label{preprocessing-steps}

@@

\subsection{Constructing the 2D model}\label{sec-construct2d}

@@

\begin{equation}\phantomsection\label{eq-equation1}{
 ar = \frac{r_2}{r_1}
}\end{equation}

\begin{equation}\phantomsection\label{eq-equation2}{
 hr = \frac{hb}{wb}
}\end{equation}

\begin{equation}\phantomsection\label{eq-equation3}{
 y_{max} = \left\lceil\frac{ar}{hr}\right\rceil * hr
}\end{equation}

@@

\begin{equation}\phantomsection\label{eq-equation11}{
 h = \sqrt{3} * s
}\end{equation}

\begin{equation}\phantomsection\label{eq-equation12}{
 b_1 = \frac{r_1 + buffer_{x}}{h}
}\end{equation}

\begin{equation}\phantomsection\label{eq-equation13}{
  v = 1.5 * s
}\end{equation}

\begin{equation}\phantomsection\label{eq-equation14}{
 b_2 = \frac{r_2 + buffer_{y}}{v}
}\end{equation}

@@

\begin{equation}\phantomsection\label{eq-equation15}{
d_x = \frac{h}{2}
}\end{equation}

\begin{equation}\phantomsection\label{eq-equation16}{
d_y = \frac{2* v * 1.15}{\sqrt{3}} 
}\end{equation}

@@

\centering{

\includegraphics[width=1\textwidth,height=1\textheight]{figures/hex_coord.png}

@@

\end{figure}%

@@

\begin{equation}\phantomsection\label{eq-equation4}{
C_k^{(2)} = (C_{ky_1}, C_{ky_2}) = \left(\frac{\sum_{i=1}^{6} h^{k}x_{i}}{6}, \frac{\sum_{i=1}^{6} h^{k}y_{i}}{6}\right).
}\end{equation}

@@

\begin{equation}\phantomsection\label{eq-equation5}{
C_k^{(2)} = (C_{ky_1}, C_{ky_2}) = \left(\frac{1}{n_k} \sum_{i=1}^{n_k} y_{1i}, \frac{1}{n_k} \sum_{i=1}^{n_k} y_{2i}\right),
}\end{equation}

@@

\subsection{Lifting the model into high
dimensions}\label{lifting-the-model-into-high-dimensions}

\subsubsection{Lifting the triangular mesh points into high
dimensions}\label{lifting-the-triangular-mesh-points-into-high-dimensions}

@@

\begin{equation}\phantomsection\label{eq-equation6}{
C_k^{(p)} = (C_{kx_1}, ..., C_{kx_p}) = \left(\frac{1}{n_k} \sum_{i=1}^{n_k} x_{1i}, \frac{1}{n_k} \sum_{i=1}^{n_k} x_{2i}, \dots ,\frac{1}{n_k} \sum_{i=1}^{n_k} x_{pi}\right).
}\end{equation}

@@

\begin{equation}\phantomsection\label{eq-equation7}{
v(C_k^{(2)}) = C_k^{(p)}.
}\end{equation}

@@

\subsubsection{Lifting the 2D triangular mesh into high
dimensions}\label{lifting-the-2d-triangular-mesh-into-high-dimensions}

@@

\centering{

\includegraphics{paper_files/figure-pdf/fig-wkhighD-1.pdf}

@@

\end{figure}%

\subsection{Tuning the model}\label{tuning-the-model}

@@

\subsubsection{Total number of bins}\label{total-number-of-bins}

@@

\centering{

\includegraphics{paper_files/figure-pdf/fig-binsize-1.pdf}

@@

\end{figure}%

\begin{equation}\phantomsection\label{eq-equation8}{
b = b_1 \times b_2
}\end{equation}

@@

\centering{

\includegraphics{paper_files/figure-pdf/fig-diagnosticpltScurve-1.pdf}

@@

\end{figure}%

\subsubsection{Benchmark value to remove low-density
hexagons}\label{benchmark-value-to-remove-low-density-hexagons}

@@

\begin{equation}\phantomsection\label{eq-equationp2}{
\text{standard count} = \frac{\text{count}}{\text{max count}} 
}\end{equation}

@@

\centering{

\includegraphics{paper_files/figure-pdf/fig-stdctsScurve-1.pdf}

@@

\centering{

\includegraphics{paper_files/figure-pdf/fig-diagnosticpltScurvelwd-1.pdf}

@@

\centering{

\includegraphics{paper_files/figure-pdf/fig-rmlowdenshex-1.pdf}

@@

\end{figure}%

\subsubsection{Benchmark value for removing long
edges}\label{benchmark-value-for-removing-long-edges}

@@

\centering{

\includegraphics{paper_files/figure-pdf/fig-distScurve-1.pdf}

@@

\centering{

\includegraphics{paper_files/figure-pdf/fig-modelScurve-1.pdf}

@@

\end{figure}%

\subsubsection{Starting point of the hexagonal
grid}\label{starting-point-of-the-hexagonal-grid}

@@

\centering{

\includegraphics{paper_files/figure-pdf/fig-mseScurvestartpoint-1.pdf}

@@

\centering{

\includegraphics{paper_files/figure-pdf/fig-scurveshifthexgridsexp-1.pdf}

@@

\centering{

\includegraphics{paper_files/figure-pdf/fig-scurveshifthexgrids-1.pdf}

@@

\end{figure}%

\subsection{Model summaries}\label{sec-summary}

\subsubsection{Predicted values and
residuals}\label{predicted-values-and-residuals}

@@

\begin{equation}\phantomsection\label{eq-equation10}{
\text{Error} = \sum_{j = 1}^{n}\sum_{i = 1}^{p} (x_{ij} - C_{x_ij})^2
}\end{equation}

@@

\subsubsection{Goodness of fit statistics}\label{sec-goodfit}

@@

\begin{equation}\phantomsection\label{eq-equation9}{
\text{MSE} = \sum_{j = 1}^{n} \frac{\sum_{i = 1}^{p} (x_{ij} - C_{x_ij})^2}{n}
}\end{equation}

\subsection{Simulated data example}\label{sec-simpleex}

@@

\centering{

\includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-nldervis5Gau-1.pdf}

@@

\end{figure}%

To investigate which is the best representation to visualize the
original data or all NLDR methods provide equivalent information, we
visualize the model-in-data space. Models from all NLDR methods show
five well-separated clusters (see Figure~\ref{fig-gau1-sc},
Figure~\ref{fig-gau2-sc}, Figure~\ref{fig-gau3-sc},
Figure~\ref{fig-gau4-sc}, and Figure~\ref{fig-gau5-sc}). This suggests
that for the five Gaussian cluster dataset, all NLDR methods effectively
preserve the global structure. tSNE, UMAP, and PHATE display clusters
with varying densities, indicating their ability to capture
within-cluster variation (see Figure~\ref{fig-gau1-sc},
Figure~\ref{fig-gau2-sc}, and Figure~\ref{fig-gau3-sc}). On the other
hand, both TriMAP and PaCMAP show clusters with flat surfaces,
suggesting a failure to capture within-cluster variation (see
Figure~\ref{fig-gau4-sc} and Figure~\ref{fig-gau5-sc}). Therefore,
TriMAP and PaCMAP do not capture the local structure as effectively as
other methods.

@@

\centering{

\includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-modelfiveGau-1.pdf}

@@

\begin{minipage}{0.33\linewidth}

\centering{

\captionsetup{labelsep=none}\includegraphics{figures/five_gau_clusters/sc_tsne_1.png}

@@

\subcaption{\label{fig-gau1-sc1}}

\end{minipage}%
%
\begin{minipage}{0.33\linewidth}

\centering{

\captionsetup{labelsep=none}\includegraphics{figures/five_gau_clusters/sc_tsne_2.png}

@@

\subcaption{\label{fig-gau1-sc2}}

\end{minipage}%
%
\begin{minipage}{0.33\linewidth}

\centering{

\captionsetup{labelsep=none}\includegraphics{figures/five_gau_clusters/sc_tsne_3.png}

@@

\subcaption{\label{fig-gau1-sc3}}

\end{minipage}%

\caption{\label{fig-gau1-sc}Screen shots of the \textbf{langevitour} of
the five Gaussian clusters dataset, shows the model with tSNE in high-D,
a video of the tour animation is available at
(\url{https://youtu.be/oQxEb4wRdHI}).}

\end{figure}%

\begin{minipage}{0.33\linewidth}

\centering{

\captionsetup{labelsep=none}\includegraphics{figures/five_gau_clusters/sc_umap_1.png}

@@

\subcaption{\label{fig-gau2-sc1}}

\end{minipage}%
%
\begin{minipage}{0.33\linewidth}

\centering{

\captionsetup{labelsep=none}\includegraphics{figures/five_gau_clusters/sc_umap_2.png}

@@

\subcaption{\label{fig-gau2-sc2}}

\end{minipage}%
%
\begin{minipage}{0.33\linewidth}

\centering{

\captionsetup{labelsep=none}\includegraphics{figures/five_gau_clusters/sc_umap_3.png}

@@

\subcaption{\label{fig-gau2-sc3}}

\end{minipage}%

\caption{\label{fig-gau2-sc}Screen shots of the \textbf{langevitour} of
the five Gaussian clusters dataset, shows the model with UMAP in high-D,
a video of the tour animation is available at
(\url{https://youtu.be/JW49csPpDx4}).}

\end{figure}%

@@

\begin{minipage}{0.33\linewidth}

\centering{

\captionsetup{labelsep=none}\includegraphics{figures/five_gau_clusters/sc_phate_1.png}

@@

\subcaption{\label{fig-gau3-sc1}}

\end{minipage}%
%
\begin{minipage}{0.33\linewidth}

\centering{

\captionsetup{labelsep=none}\includegraphics{figures/five_gau_clusters/sc_phate_2.png}

@@

\subcaption{\label{fig-gau3-sc2}}

\end{minipage}%
%
\begin{minipage}{0.33\linewidth}

\centering{

\captionsetup{labelsep=none}\includegraphics{figures/five_gau_clusters/sc_phate_3.png}

@@

\subcaption{\label{fig-gau3-sc3}}

\end{minipage}%

\caption{\label{fig-gau3-sc}Screen shots of the \textbf{langevitour} of
the five Gaussian clusters dataset, shows the model with PHATE in
high-D, a video of the tour animation is available at
(\url{https://youtu.be/xqhRODt8Z-o}).}

\end{figure}%

\begin{minipage}{0.33\linewidth}

\centering{

\captionsetup{labelsep=none}\includegraphics{figures/five_gau_clusters/sc_trimap_1.png}

@@

\subcaption{\label{fig-gau4-sc1}}

\end{minipage}%
%
\begin{minipage}{0.33\linewidth}

\centering{

\captionsetup{labelsep=none}\includegraphics{figures/five_gau_clusters/sc_trimap_2.png}

@@

\subcaption{\label{fig-gau4-sc2}}

\end{minipage}%
%
\begin{minipage}{0.33\linewidth}

\centering{

\captionsetup{labelsep=none}\includegraphics{figures/five_gau_clusters/sc_trimap_3.png}

@@

\subcaption{\label{fig-gau4-sc3}}

\end{minipage}%

\caption{\label{fig-gau4-sc}Screen shots of the \textbf{langevitour} of
the five Gaussian clusters dataset, shows the model with TriMAP in
high-D, a video of the tour animation is available at
(\url{https://youtu.be/X-uHSB4VoZ0}).}

\end{figure}%

@@

\begin{minipage}{0.33\linewidth}

\centering{

\captionsetup{labelsep=none}\includegraphics{figures/five_gau_clusters/sc_pacmap_1.png}

@@

\subcaption{\label{fig-gau5-sc1}}

\end{minipage}%
%
\begin{minipage}{0.33\linewidth}

\centering{

\captionsetup{labelsep=none}\includegraphics{figures/five_gau_clusters/sc_pacmap_2.png}

@@

\subcaption{\label{fig-gau5-sc2}}

\end{minipage}%
%
\begin{minipage}{0.33\linewidth}

\centering{

\captionsetup{labelsep=none}\includegraphics{figures/five_gau_clusters/sc_pacmap_3.png}

@@

\subcaption{\label{fig-gau5-sc3}}

\end{minipage}%

\caption{\label{fig-gau5-sc}Screen shots of the \textbf{langevitour} of
the five Gaussian clusters dataset, shows the model with PaCMAP in
high-D, a video of the tour animation is available at
(\url{https://youtu.be/UxvU8HQd7JM}).}

\end{figure}%

\section{Applications}\label{sec-applications}

\subsection{Single-cell RNA-seq data of
human}\label{single-cell-rna-seq-data-of-human}

@@

\centering{

\includegraphics{paper_files/figure-pdf/fig-umapauthor-1.pdf}

@@

\end{figure}%

To determine whether the UMAP representation with the parameter choice
suggested by \citet{chen2023} preserves the original data structure, we
visualize the model constructed with UMAP overlaid on the high-D data
(model-in-data space). The figures in Figure~\ref{fig-pbmc1-sc} show
three well-separated clusters, indicating that the suggested UMAP
representation preserves the global structure (see
Figure~\ref{fig-umapauthor}). However, as shown in
Figure~\ref{fig-pbmc1-sc}, these clusters are close to each other in
high-D space. Also, nonlinear continuity patterns and high-density
patches within the clusters are observed (see
Figure~\ref{fig-pbmc1-sc}). Therefore, the suggested UMAP representation
(see Figure~\ref{fig-umapauthor}) does not accurately preserve the local
structure of the PBMC3k dataset.
